---
title: "Exploring the Effect of Elevation & Season on Avalanche Accidents"
author: Nicole Pepper
format: 
  html:
    toc: true
editor_options: 
  chunk_output_type: console
output: html_document

---

# Set up workspace
Load packages
```{r}
# ---- Load required packages ----
library(tidyverse)
library(ggplot2)
library(here)
library(knitr)
library(kableExtra)
library(terra)
library(sf)
library(elevatr)
```

#### Read in Data
```{r}
# ---- Read in data ----
av_accidents <- read_csv(here::here("data","avalanche_accidents.csv")) 

```

```{r}
# ---- Prepare av_accidents data ----

# Clean lat and lon columns by removing NAs, 0, and #REF values
av_accidents <- av_accidents %>% 
    drop_na(lat, lon) %>% # Drop NAs in lat and lon columns
    mutate(across(c(lat, lon), as.character)) %>%  # Ensure columns are characters for filtering
    filter(lat != "0", lon != "0", lat != "#REF!", lon != "#REF!") %>% # Remove values that equal 0 or #REF!
    mutate(across(c(lat, lon), as.numeric))  # Convert back to numeric

# Convert avalanche accidents to spatial object
av_points <- sf::st_as_sf(av_accidents, coords = c("lon", "lat"), crs = 4326)
```

```{r}
plot(av_points)
```

# Extract elevation at accidents
```{r}

# ---- Extract elevation at avalanche accident locations ----
# Fetch elevation data
av_points <- get_elev_point(locations = av_points, prj = sf::st_crs(4326)$proj4string)

# Convert the result back to a data frame by replacing av_accidents
av_accidents <- as.data.frame(av_points)
```

# Correlation Matrix
```{r}

```

# Visualize accidents by elevation
```{r}

```

# Randomization test for elevation 
```{r}
# ---- Point estimate success for observed avalanche accidents ----

# Calculate conflict success for observed data

# Calculate the difference in []

# ---- Calculate the null distribution for avalanche accidents ----

# ---- Visualize the new proportions of randomized data for null distribution ----


# ---- Calculate p value  ----

```


# Visualize accidents by month
```{r}

```

# Fit a logistic model:

### Avalanche accidents ~ elevation + season + elevation:season

```{r}
# ---- Fit logistic regression model to avalanche accident data ----


# Get coefficients

```

Examine the p-values for elevation, season, & elevation:season


# Visualize model results

```{r}
# Plot the predicted avalanche accidents by elevation and season

```

Optional: CAIC wants to focus their efforts on avalanche risk education and outreach. Which elevation/zone months should they focus on incidents of  on 20% fatalities...
```{r}

```

