---
title: "Exploring the Effect of Elevation & Season on Avalanche Accidents"
author: Nicole Pepper
format: 
  html:
    toc: true
editor_options: 
  code-fold: true
  chunk_output_type: console
output: html_document
---


Overview: 

In this assignment, I am interested in the relationship between month and elevation on avalanche accidents in Colorado.


Null Hypothesis (H₀): There is no interaction between elevation zones and months in determining avalanche fatality rates. Specifically, the avalanche fatality rates do not differ between elevation zones across months.

This phrasing highlights the interaction you're testing, clarifies the levels of elevation, and connects them explicitly to the variable of time (months). - Alternative Hypothesis: Avalanche fatality rates differ by elevation zone and month
# 1. Set up workspace

### Load packages
```{r}
# ---- Load required packages ----

library(tidyverse)
library(ggplot2)
library(here)
library(knitr)
library(kableExtra)
library(terra)
library(sf)
library(elevatr)
library(hexbin)
```

# 2. Read in Project Data
```{r}
# ---- Read in data ----
av_accidents <- read_csv(here::here("data","avalanche_accidents.csv")) %>%
  filter(State == "CO") # Filter for Colorado (CO) in State Column

```

# 3. Prepare Data

### Prepare Avalanche Accidents Data
```{r}
# ---- Clean av_accidents data ----

# Clean lat and lon columns by removing NAs, 0, and #REF values
av_accidents <- av_accidents %>% 
    drop_na(lat, lon) %>% # Drop NAs in lat and lon columns
    mutate(across(c(lat, lon), as.character)) %>%  # Ensure columns are characters for filtering
    filter(lat != "0", lon != "0", lat != "#REF!", lon != "#REF!") %>% # Remove values that equal 0 or #REF!
    mutate(across(c(lat, lon), as.numeric))  # Convert back to numeric

# Convert avalanche accidents to spatial object to extract elevation at geometry
av_points <- sf::st_as_sf(av_accidents, coords = c("lon", "lat"), crs = 4326)
```

### Extract Elevation at Accident Locations
```{r}
# ---- Extract elevation at avalanche accident locations ----

# Call in elevation data using `elevatr`
av_points <- get_elev_point(locations = av_points,
                            prj = sf::st_crs(4326)$proj4string)

# Convert the result back to a data frame by replacing av_accidents
av_accidents <- as.data.frame(av_points)

# ---- Clean and remove unwanted elevation values ---- 

av_accidents <- av_accidents %>% 
    filter(elevation != 0, elevation != -10000) 

# ---- Add a column for elevation band  ---- 

# Define and add a column for elevation groups
av_accidents <- av_accidents %>%
  mutate(elevation_group = case_when(
    elevation < 3352.8 ~ "Below Treeline",
    elevation >= 3352.8 & elevation <= 3652 ~ "Near Treeline",
    elevation > 3652 ~ "Above Treeline"
  ))

```

### Prepare Month Column
```{r}
# ---- Prepare month column for preliminary plots ----

# Define list of Month Names for plot labels
month_names <- c(
  "1" = "January",
  "2" = "February",
  "3" = "March",
  "4" = "April",
  "5" = "May",
  "6" = "June",
  "7" = "July",
  "9" = "September",
  "10" = "October",
  "11" = "November",
  "12" = "December"
)

# Define av_accidents_plot data to be used for the plot, prepping this separately because I don't want to use this for the glm
av_accidents_plot <- av_accidents

# Define and apply levels for winter season
av_accidents_plot$MM <- factor(av_accidents$MM,
                          levels = c(9, 10, 11, 12, 1, 2, 3, 4, 5, 6, 7, 8),
                          ordered = TRUE)

# Map MM column values to their corresponding names and order for plot
av_accidents_plot$month_names <- factor(
  av_accidents$MM,
  levels = c(11, 12, 1, 2, 3, 4, 5, 6, 7, 9, 10),
  labels = c("November", "December", "January", "February", "March", "April", 
             "May", "June", "July", "September", "October"),
  ordered = TRUE
)

# ---- Prepare month column for glm ----

# Add a column for distance from March
av_accidents <- av_accidents %>%
  mutate(
    MM = factor(MM, levels = c(9, 10, 11, 12, 1, 2, 3, 4, 5, 6, 7, 8), ordered = TRUE),
    dist_from_march = as.numeric(MM) - which(levels(MM) == 3)
  )

# Add a column for distance from March
av_accidents <- av_accidents %>%
  mutate(
    dist_from_march = as.numeric(MM) - match(3, levels(MM))
  )

```

### Filter data to Colorado

# 4. Visualize Preliminary Data

### Visualize Accidents by Elevation

```{r}
# ---- Plot accidents by elevation group ----

ggplot(data = av_accidents) + 
  geom_histogram(aes(x = elevation),
                fill = "#4ECBDE",
                alpha = 0.8) +
geom_vline(xintercept = c(3352.8, 3652), color = "firebrick", size = 0.75, linetype = "solid") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16),
       axis.title.x = element_text(angle = 0,
                                   vjust = 0.5,
                                   face = "bold"),
       axis.title.y = element_text(angle = 0,
                                   vjust = 0.5,
                                   face = "bold")) +
  labs(title = "Avalanches Accidents by Elevation Zone in Colorado",
       x = "Elevation (Meters)",
       y = "Avalanche\nAccidents") +
  geom_text(aes(x = 3100, y = .5), 
           label = "Below Treeline", size = 4, fontface = "italic") +
  geom_text(aes(x = 3500, y = .5), 
           label = "Near Treeline", size = 4, fontface = "italic") +
  geom_text(aes(x = 3850, y = .5), 
           label = "Above Treeline", size = 4, fontface = "italic")
```


```{r}
# ---- Plot accidents by month ----
ggplot(data = av_accidents) + 
  geom_bar(aes(x = month_names),
                fill = "#4ECBDE",
                alpha = 0.8) +
  scale_x_discrete(labels = c(
    "1" = "January",
    "2" = "February",
    "3" = "March",
    "4" = "April",
    "5" = "May",
    "6" = "June",
    "7" = "July",
    "9" = "September",
    "10" = "October",
    "11" = "November",
    "12" = "December"
  )) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold",size = 16),
        axis.text.x = element_text(angle = 0,
                                   hjust = 0.5,
                                   face = "bold"),
        axis.title.y = element_text(angle = 0,
                                    vjust = 0.5,
                                    face = "bold"),
        axis.title.x = element_text(angle = 0,
                                    face = "bold")) +
  geom_text(
    aes(x = factor(month_names), label = ..count..),
    stat = "count",                        
    vjust = -0.5,
    color = "firebrick",
    fontface = "bold"
  ) +
  labs(title = "Avalanche Accidents in Colorado by Month",
       x = "Month",
       y = "Fatal\nAccidents")
```

# ---- Randomization test for month ----

# Fit a logistic model:

## *How does month and elevation impact the probability of a fatal avalanche accident?*
When they put out their reports they have a risk index for below, at, and above treeline.

### Avalanche accidents \~ elevation + season + elevation:season
```{r}
# Practice

# Aggregate data by Month and elevation group
av_accidents_agg <- co_av_accidents %>%
  group_by(dist_, elevation_group) %>%
  summarise(accidents = n(), .groups = "drop") %>%
  arrange(month_names, elevation_group) # Optional: arrange for better viewing


# Fit the avalanche accident model
av_accident_glm <- glm(accidents ~ elevation_group + month_names + elevation_group:month_names, 
             family = poisson(link = "log"), 
             data = co_av_accidents_agg)


summary(av_accident_glm)
```


```{r}
# Aggregate data by Month and elevation group
av_accidents_agg_march <- av_accidents %>%
  group_by(elevation_group, dist_from_march) %>%
  summarise(accidents = n(), .groups = "drop") %>%
  arrange(elevation_group, dist_from_march) # Optional: arrange for better viewing

# Fit the GLM
avalanche_glm_march <- glm(
  formula = accidents ~ elevation_group + dist_from_march + elevation_group:dist_from_march,
  data = av_accidents_agg_march,
  family = poisson(link = "log")  # Assuming count data follows a Poisson distribution
)

# Summarize the model
summary(avalanche_glm_march)
```

**Poisson(link = "log")**: is the standard family for modeling count data, which assumes that the response variable follows a Poisson distribution with a mean that is a function of the predictors. The `log` link is commonly used because it ensures the predicted count remains positive.

Residuals after 4000 m are very negative, likely because not as many people recreate above that elevation.


Examine the p-values for elevation, season, & elevation:season

# Visualize model results

```{r}
# Plot the predicted avalanche accidents by elevation and season

```

Optional: CAIC wants to focus their efforts on avalanche risk education and outreach. Which elevation/zone months should they focus on incidents of on 20% fatalities...

